Accelerating structure–function mapping using the ViVa webtool to mine natural variation

### 2018-11-21 manuscript/01-intro.Rmd
updates from JLN 181118

The first genome sequence of *Arabidopsis thaliana* facilitated rapid advancement of plant biology [@thearabidopsisgenomeinitiative_analysis_2000]. Since this initial genome, massive scaling of sequencing technology has allowed the survey the genomic variation in natural *Arabidopsis thaliana* populations [@nordborg_pattern_2005; @borevitz_genomewide_2007; @weigel_1001_2009]. This valuable population genetics resource has led to several associations of genetic loci with phenotypic traits and provided insights into how selective pressure has influenced the evolution of plant genomes [@long_massive_2013; @atwell_genomewide_2010; @clark_common_2007]. Despite the high value of this compendium of genetic variation, its use has been largely limited to investigations of population genetics and for genome-wide association studies.

Natural genetic variation provides a catalog of permissible polymorphisms which may facilitate exploration and prediction of genotype/function/phenotype relationships at the gene, gene family and network scales [@joly-lopez_developing_2016]. Outside of large effect variants, these relationships are difficult to map. Such studies in plants have the added obstacles of large gene families and the still quite time and resource-intensive process of allele replacement. Massively parallel assays of variant effects have revolutionized genetic diagnostics and personalized medicine [@starita_variant_2017; @gasperini_power_2016; @matreyek_platform_2017]. We envision the use of plant natural variation datasets as a tool to similarly revolutionize breeding and genetic engineering of crop plants by rapidly advancing our understanding of genotype/function/phenotype relationships. Biochemical or synthetic assays of gene function and quantitative phenotyping of the germplasm of collections of natural variation can facilitate functional and phenotypic quantification [@wright_insights_2017]. However, the high dimensionality of natural variation datasets often limits access to those with bioinformatics expertise.

Here, we present ViVa: a web application and R-package for Visualizing Variation, which allows plant molecular biologists of any level access to the 1001 genomes database. ViVa facilitates formulation of hypotheses as to the genotype/function/phenotype relationships within any gene, family, or network. Using ViVa researchers may: 1. Identify polymorphisms to facilitate biochemical assays of variant effects [@starita_variant_2017; @wright_insights_2017]; 2. Produce family-wise alignments of variants to facilitate de novo functional domain identification [@melamed_combining_2015]; 3. Generate lists of accessions containing polymorphisms to facilitate phenotypic analysis of gene variant effects [@park_distributions_2017]; and 4. Quantify metrics of genetic diversity to facilitate the study of gene, gene family and network evolution [@delker_natural_2010; @kliebenstein_role_2008].

>>>>

The sequencing of the first *Arabidopsis thaliana* genome ushered in a new era of tool development and systematic functional annotation of plant genes [@thearabidopsisgenomeinitiative_analysis_2000]. Since that landmark effort, massive scaling of sequencing technology has allowed for the survey of genomic variation in natural *Arabidopsis thaliana* populations [@nordborg_pattern_2005; @borevitz_genomewide_2007; @weigel_1001_2009]. This valuable population genetics resource has led to several associations of genetic loci with phenotypic traits and provided insights into how selective pressure has influenced the evolution of plant genomes [@long_massive_2013; @atwell_genomewide_2010; @clark_common_2007].

Beyond its utility in gene discovery and understanding genome evolution, natural genetic variation provides a catalog of permissible polymorphisms that can facilitate the connection  of genotype to phenotype at the gene, gene family and network scales [@joly-lopez_developing_2016]. This is an espcecially critica resource in large gene families where loss of function in individual genes may have little or no phenotypcial effect, and directed allele replacement remains time and resource-intensive. Massively parallel assays of variant effects have revolutionized genetic diagnostics and personalized medicine [@starita_variant_2017; @gasperini_power_2016; @matreyek_platform_2017]. We envision the use of plant natural variation datasets as a tool to similarly revolutionize breeding and genetic engineering of crop plants by rapidly advancing our understanding of genotype/function/phenotype relationships. A proof-of-principle survey of a relatively small subset of natural variants paired with a synthetic assay of gene function successfully mapped critical functional domains of auxin receptors [@wright_insights_2017].

So why is the survey of natural variants not as routine as a BLAST search? One reason may be the current requirement for a fairly high level of bioinformatics expertise to extract the desired information from the whole genome resequencing datasets. To address this concern, we created ViVa: a web application and R-package for Visualizing Variation, which allows plant molecular biologists of any level access to gene level data from the 1001 genomes database. Using ViVa researchers may: 1) Identify polymorphisms to facilitate biochemical assays of variant effects [@starita_variant_2017; @wright_insights_2017]; 2) Produce family-wise alignments of variants to facilitate de novo functional domain identification [@melamed_combining_2015]; 3) Generate lists of accessions containing polymorphisms to facilitate phenotypic analysis of gene variant effects [@park_distributions_2017]; and 4) Quantify metrics of genetic diversity to facilitate the study of gene, gene family and network evolution [@delker_natural_2010; @kliebenstein_role_2008].

### 2018-11-21 manuscript/02-results.Rmd
updates from JLN 181118

The current version of ViVa is specifically focused on visualizes natural variation in the coding sequences of genes. Future versions will expand upon this to examine noncoding sequences; however, indels are much more common in noncoding sequences, which limits the certainty of this resequencing data.

**Morgan will provide screenshots**
### Data input: Select the desired genes

@@

To test the useability and accessibility of this tool we assembled a group of alpha testers comprising postdoctoral, graduate, and undergraduate researchers. This team of researchers analyzed natural variation in the nuclear auxin signaling pathway. This signaling pathway has been well studied, revealing abundant functional knowledge and crystal structures of several protein domains, providing several points of ground truth reference for comparing structure-function variation and natural genetic variation.

Assuming there is selective pressure against mutations which alter a protein's function, nonsynonymous genetic variation in critical functional domains of a protein should be low. Therefore scanning gene coding sequences for regions of low nonsynonymous diversity, may predict functional domains. Within the auxin signaling pathway several domains have been identified as critical to transmitting auxin signals.

Perhaps the most critical domain is the degron of the Aux/IAA transcriptional repressor proteins. Mutations which prevent this domain from being recognized and marked for proteasomal degradation by SCF^TIR1/AFB^ are dominant and often result in obvious phenotypes [@berleth_auxin_2004; @yang_iaa1_2004]. As expected this critical functional region is highly conserved at the amino acid level (\@ref(fig:A-IAA-alignment)). Similarly, the EAR motif which facilitates Aux/IAA interaction with TPL/TPR transcriptional repressors, are also conserved.

@@

(ref:A-IAA-alignment) **Canonical IAAs have conserved Degron and EAR motifs.** Protein sequences were aligned [@wright_decipher_2015] and variants were mapped to this alignment and colored according to the predicted functional effect of the most deleterious variant at that position, with light colors being less deleterious and darker colors being more deleterious. Red indicates missense variants. Color scale is explained in [Methods](#methods). EAR motif is highlighted by a black box including alignment positions 70-74. Degron domain is highlighted by a gray box including alignment positions 194-201.
```{r A-IAA-alignment, fig.cap="(ref:A-IAA-alignment)"}
A_IAAs <- subset(gene_info, tair_symbol %in% IAA_data[which(IAA_data$Class == "A"), "name"])
A_IAAs <- alignCDS(A_IAAs$transcript_ID)
# make an alignment data frame
A_IAAs <- makeAlnDF(A_IAAs[[2]])
A_IAAs <- left_join(A_IAAs, dplyr::select(gene_info, "tair_locus",
                                       "tair_symbol", "transcript_ID"),
                        by = c("transcript_ID" = "transcript_ID"))
A_IAAs$tair_symbol %<>% as.factor()
A_IAAs$tair_symbol %<>% fct_reorder(
    {as.integer(str_remove(A_IAAs$tair_symbol, 
                           pattern = "IAA"))}, 
    .desc = TRUE)
VCF <- ldply(.data = VCF, .fun = subset, gt_GT != "0|0")
coding_VCF <- getCodingDiv(VCF)
A_IAAs <- addSNPsToAlnDF(aln_df = A_IAAs,SNPs = coding_VCF, effect_order = SNPeff_order)
A_IAAs$gap <- A_IAAs$seq_pos == "-"

p <-ggplot(A_IAAs, aes(x = aln_pos, y = tair_symbol)) +
  geom_raster(data = A_IAAs, 
             # mapping = aes(fill = effects)) +
              mapping = aes(fill = strength,
                            alpha = gap)) +
  scale_alpha_manual(values = c("TRUE" = 0, "FALSE" = 1), 
                     labels = c("aligned", "gap"), 
                     guide = guide_legend(override.aes = 
                                            list(fill = c("grey85", "white")),
                       title = "alignment",
                       direction = "vertical")) + 
   scale_fill_viridis(breaks = range(SNPeff_order$strength),
                      limits = range(SNPeff_order$strength),
                      na.value = "grey85",
                      direction = -1,
                      labels = c("neutral", "deleterious"),
                      option = "A", 
                      guide = guide_colorbar(ticks = FALSE, 
                                             title.vjust = .8, 
                                             title.position = "top", 
                                             title.hjust = 0.5)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_x_continuous(expand = c(0,0)) +  # expand increases distance from axis
  labs(x = "codon position in alignment", y = "", fill = "variant effect") +
  theme_logo(base_family = "Helvetica") +
  theme(panel.grid = element_blank(), 
        panel.grid.minor = element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_blank(),
        #panel.background = element_rect(fill = "grey85", color = "white"),
        legend.position = c("bottom"),
        legend.key = element_rect(linetype = "solid"),
        legend.spacing = unit(4, "char"), 
        legend.box.spacing = unit(0.1,"line")) 
p + geom_rect(mapping = aes(xmin = 193, xmax = 202, ymin = 0, ymax = length(levels((tair_symbol)))+1), color= "black", fill = NA) + geom_rect(mapping = aes(xmin = 69, xmax = 75, ymin = 0, ymax = length(levels((tair_symbol)))+1), color= "black", fill = NA)
```

The Aux/IAA proteins bind to ARF transcription factors allowing the Aux/IAA's to effect transcriptional repression at specific loci. Thus in the ARF transcription factor proteins we expect the DNA-binding function to be conserved. Indead, the DNA-binding domains of the ARFs had very few, low-diversity missense polymorphisms, as did the PB1 domain which facilitates ARF-Aux/IAA interaction (Figure \@ref(fig:A-ARF-Alignment)).

```{r ARF diversity stats}
gene_info <- geneInfoFromFile(
  fname = system.file("shiny-app", "gene_family_data",
                      "ARF_gene_ids.csv", package = "r1001genomes"))

VCF <- readRDS(system.file("shiny-app", "gene_family_data",
                      "ARF_VCF.rds", package = "r1001genomes"))

VCF <- ldply(.data = VCF, .fun = subset, gt_GT != "0|0")
VCF <- getCodingDiv(VCF)
```

```{r Mask_ARF_alignment, eval=FALSE}
aliscore(as.AAbin(alignment[[2]]), path = "/Volumes/SSD/Applications/ALISCORE_v2.0/Aliscore_v.2.0/")
id <- scan(file = "/Volumes/SSD/Applications/ALISCORE_v2.0/Aliscore_v.2.0/input.fas_List_random.txt", sep = " ", quiet = TRUE)
AAAlign_no_gaps <- as.matrix(alignment[[2]])[, id]
write.nexus.data(AAAlign_no_gaps, file = "ARFs.nex", format = "protein")
```
```{bash ARF_phylogeny_construction, eval=FALSE}
/Applications/MrBayes/mb

execute ARFs.nex
# set ARF17 as the outgroup
outgroup AT1G77850.1

```

```{r Load ARF phylogeny and data}
tree <- treeio::read.beast(file = "trees/ARFs.nex.con.tre")

ARF_data <- read.csv(system.file("shiny-app", "gene_family_data",
                      "ARF_gene_ids.csv", package = "r1001genomes"))

#tip_labels need to be in the first column
ARF_data %<>% dplyr::select(name, everything())
ARF_data$tair_locus %<>% as.character()

ARF_data <- ARF_data[match(gsub(x = tree@phylo$tip.label, pattern = "\\.1",replacement = ""), ARF_data$tair_locus), ]

tree@phylo$tip.label <- ARF_data$name

```

(ref:A-ARF-Alignment) **Canonical ARFs have conserved DNA binding and PB1 domains.** Protein sequences were aligned [@wright_decipher_2015] and variants were mapped to this alignment and colored according to the predicted functional effect of the most deleterious variant at that position, with light colors being less deleterious and darker colors being more deleterious. Red indicates missense variants. Color scale is explained in [Methods](#methods).
```{r A-ARF-Alignment, fig.cap="(ref:A-ARF-Alignment)"}
A_ARFs <- subset(gene_info, tair_symbol %in% c("ARF5", "ARF6", "ARF7", "ARF8", "ARF19"))
A_ARFs <- alignCDS(A_ARFs$transcript_ID)
# make an alignment data frame
A_ARFs <- makeAlnDF(A_ARFs[[2]])

A_ARFs <- left_join(A_ARFs, dplyr::select(gene_info, "tair_locus",
                                       "tair_symbol", "transcript_ID"),
                        by = c("transcript_ID" = "transcript_ID"))
#A_ARFs$tair_symbol %<>% factor(levels = ARF_order, 
#                                       ordered = TRUE)
A_ARFs$tair_symbol %<>% fct_rev()

A_ARFs <- addSNPsToAlnDF(A_ARFs, VCF, effect_order = SNPeff_order)
A_ARFs$gap <- A_ARFs$seq_pos == "-"

p <-ggplot(A_ARFs, aes(x = aln_pos, y = tair_symbol)) +
  geom_raster(data = A_ARFs, 
             # mapping = aes(fill = effects)) +
              mapping = aes(fill = strength,
                            alpha = gap)) +
  scale_alpha_manual(values = c("TRUE" = 0, "FALSE" = 1), 
                     labels = c("aligned", "gap"), 
                     guide = guide_legend(override.aes = 
                                            list(fill = c("grey85", "white")),
                       title = "alignment",
                       direction = "vertical")) + 
   scale_fill_viridis(breaks = range(SNPeff_order$strength),
                      limits = range(SNPeff_order$strength),
                      na.value = "grey85",
                      direction = -1,
                      labels = c("neutral", "deleterious"),
                      option = "A", 
                      guide = guide_colorbar(ticks = FALSE, 
                                             title.vjust = .8, 
                                             title.position = "top", 
                                             title.hjust = 0.5)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_x_continuous(expand = c(0,0)) +  # expand increases distance from axis
  labs(x = "codon position in alignment", y = "", fill = "variant effect") +
  theme_logo(base_family = "Helvetica") +
  theme(panel.grid = element_blank(), 
        panel.grid.minor = element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_blank(),
        #panel.background = element_rect(fill = "grey85", color = "white"),
        legend.position = c("bottom"),
        legend.key = element_rect(linetype = "solid"),
        legend.spacing = unit(4, "char"), 
        legend.box.spacing = unit(0.1,"line")) 
p
#ggsave(p,filename = "A_ARFs_alignment_variants.pdf", width = 12, height = 4)
```

Natural variation also provides a means to study how gene families are evolving. Mapping diversity at nonsynonymous variant sites relative to synonymous sites to gene trees allows comparison of recent functional divergence to the sequence divergence throughout the growth of the gene family. This visualization facilitates hypothesis generation regarding the functional redundancy or novelty of recent gene duplications.

Previous research has found evidence of both broad genetic redundancy and specificity within closely related pairs or groups of IAA proteins [@overvoorde_functional_2005]. For example, the *iaa8-1 iaa9-1* double mutant and the *iaa5-1 iaa6-1 iaa19-1* triple mutant have wild-type phenotypes [@overvoorde_functional_2005], yet the *IAA6/19* sister pair has significant differences in expression patterns, protein abundances and functions suggesting they have undergone functional specialization since their divergence [@winkler_variation_2017].

In our analysis of the *IAA* gene family, we frequently observed that one member of most *IAA* sister pairs had high nonsynonymous diversity while the other sister was more conserved \@ref(fig:IAAtree). In an interspecific comparison between *A. thaliana* and *A. lyrata*, *IAA19* is more conserved, while *IAA6* has regions of increased sequence divergence upstream of the degron and in the N-terminal half of the PB1 domain [@winkler_variation_2017]. Consistent with these results, our analysis of intraspecific natural variation revealed higher conservation for *IAA19* (${\pi_N/\pi_S}$ = 0.55) compared to *IAA6* (${\pi_N/\pi_S}$ = 2.3) \@ref(fig:IAAtree), and also detected high diversity within the same regions of *IAA6* as seen in Winkler *et al*.

@@

Within the Group B *IAA* genes, there are four sister pairs---*IAA10*/*IAA11*, *IAA12*/*IAA13*, *IAA18*/*IAA26*, and *IAA20*/*IAA30* \@ref(fig:IAAtree) [@remington_contrasting_2004]. Interestingly, among these sister pairs only *IAA12* and *IAA13* have a noticeable difference in nonsynonymous diversity. The *Arabidopsis thaliana* EFP browser shows that *IAA10* and *IAA11* have almost identical expression patterns, with the exception of *IAA10* being more highly expressed in the developing seed, and *IAA11* being more highly expressed in root tissue during lateral root development **need to cite original data source here**. In our analysis, *IAA10* and *IAA11* both showed functional conservation (${\pi_N/\pi_S}$ of 0.80 and 0.67 respectively) in support of *IAA10* and *IAA11* playing redundant roles. Accessions and associated variants identified here provide a potentially expedited route to elucidating the extent of redundancy between family members such as *IAA10* and *IAA11*.

>>>>

ViVa is meant to visualize natural variation in the coding sequences of genes. For this first iteration, non-coding sequence variation is intentionally excluded from the analysis tools. This reflects challenges both in alignment of non-coding sequences and the increased difficulty in assessing variation in these regions [@alexandre_complex_2018].

**Morgan will provide screenshots**

### Data input: Select the desired genes

@@

To test the useability and accessibility of ViVa, we assembled a group of alpha testers comprising postdoctoral, graduate, and undergraduate researchers at a research university (University of Washington) and at a primarily-undergraduate institutions (Whitman College). Our testers focused their investigation of natural variation on the nuclear auxin signaling pathway. We selected this signaling pathway for multiple reasons including a wealth of functional data and solved structures of several domains or entire proteins. In this way, we were able to assess the predictive ability of the ViVa modules. Summary information about each family examined can be found in the supplement. Below, we describe the results for the Aux/IAA family in more detail.

In most cases, natural selection is expected to minimize the persistence of nonsynonymous mutations in sequences that encode critical functional domains  relative to their persistance in non-critical domains. Therefore, we reasoned that scanning gene coding sequences for regions of relatively low nonsynonymous diversity should highlight functional domains. This general principle can be seen clearly in the analysis of the Aux/IAA family of transcriptional co-repressors/co-receptors. Aux/IAAs have three major domains: domain I facilitates interaction with TPL/TPR transcriptional repressors, domain II facilitates interactions with the TIR1/AFB receptors in the presence of auxin, and the PB1 domain which failitates interactions with the ARF transcription factors. Domain 1 and Domain II can be readily identified (\@ref(fig:A-IAA-alignment)), perhaps because the multiple contributing residues are spread out in linear sequence space. It is worth noting that the most important residues within the PB1 domain also show very little variation.

@@

(ref:A-IAA-alignment) **Canonical IAAs have conserved Degron and EAR motifs.** Protein sequences were aligned [@wright_decipher_2015] and variants were mapped to this alignment and colored according to the predicted functional effect of the most deleterious variant at that position, with light colors being less deleterious and darker colors being more deleterious. Red indicates missense variants. Color scale is explained in [Methods](#methods). EAR motif is highlighted by a black box including alignment positions 70-74. Degron domain is highlighted by a box including alignment positions 194-201. Critical residues for PB1 domain interactions are highlighted at alignment positions 246, 253, 316, and 320.
```{r A-IAA-alignment, fig.cap="(ref:A-IAA-alignment)"}
A_IAAs <- subset(gene_info, tair_symbol %in% IAA_data[which(IAA_data$Class == "A"), "name"])
A_IAAs <- alignCDS(A_IAAs$transcript_ID)
# make an alignment data frame
A_IAAs <- makeAlnDF(A_IAAs[[2]])
A_IAAs <- left_join(A_IAAs, dplyr::select(gene_info, "tair_locus",
                                       "tair_symbol", "transcript_ID"),
                        by = c("transcript_ID" = "transcript_ID"))
A_IAAs$tair_symbol %<>% as.factor()
A_IAAs$tair_symbol %<>% fct_reorder(
    {as.integer(str_remove(A_IAAs$tair_symbol, 
                           pattern = "IAA"))}, 
    .desc = TRUE)
VCF <- ldply(.data = VCF, .fun = subset, gt_GT != "0|0")
coding_VCF <- getCodingDiv(VCF)
A_IAAs <- addSNPsToAlnDF(aln_df = A_IAAs,SNPs = coding_VCF, effect_order = SNPeff_order)
A_IAAs$gap <- A_IAAs$seq_pos == "-"

p <-ggplot(A_IAAs, aes(x = aln_pos, y = tair_symbol)) +
  geom_raster(data = A_IAAs, 
             # mapping = aes(fill = effects)) +
              mapping = aes(fill = strength,
                            alpha = gap)) +
  scale_alpha_manual(values = c("TRUE" = 0, "FALSE" = 1), 
                     labels = c("aligned", "gap"), 
                     guide = guide_legend(override.aes = 
                                            list(fill = c("grey85", "white")),
                       title = "alignment",
                       direction = "vertical")) + 
   scale_fill_viridis(breaks = range(SNPeff_order$strength),
                      limits = range(SNPeff_order$strength),
                      na.value = "grey85",
                      direction = -1,
                      labels = c("neutral", "deleterious"),
                      option = "A", 
                      guide = guide_colorbar(ticks = FALSE, 
                                             title.vjust = .8, 
                                             title.position = "top", 
                                             title.hjust = 0.5)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_x_continuous(expand = c(0,0)) +  # expand increases distance from axis
  labs(x = "codon position in alignment", y = "", fill = "variant effect") +
  theme_logo(base_family = "Helvetica") +
  theme(panel.grid = element_blank(), 
        panel.grid.minor = element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_blank(),
        #panel.background = element_rect(fill = "grey85", color = "white"),
        legend.position = c("bottom"),
        legend.key = element_rect(linetype = "solid"),
        legend.spacing = unit(4, "char"), 
        legend.box.spacing = unit(0.1,"line")) 
p + geom_rect(mapping = aes(xmin = 193, xmax = 202, ymin = 0, ymax = length(levels((tair_symbol)))+1), color= "black", fill = NA) + geom_rect(mapping = aes(xmin = 69, xmax = 75, ymin = 0, ymax = length(levels((tair_symbol)))+1), color= "black", fill = NA) + geom_rect(mapping = aes(xmin = 245, xmax = 247, ymin = 0, ymax = length(levels((tair_symbol)))+1), color = "black", fill = NA, size = .2) +
geom_rect(mapping = aes(xmin = 319, xmax = 321, ymin = 0, ymax = length(levels((tair_symbol)))+1), color = "black", fill = NA, size = 0.2) +
  geom_rect(mapping = aes(xmin = 252, xmax = 254, ymin = 0, ymax = length(levels((tair_symbol)))+1), color = "black", fill = NA, size = 0.2) +
  geom_rect(mapping = aes(xmin = 315, xmax = 317, ymin = 0, ymax = length(levels((tair_symbol)))+1), color = "black", fill = NA, size = 0.2) 
```

Natural variation also provides a means to study how gene families are evolving. To do this, we used ViVa to map the diversity at nonsynonymous variant sites relative to synonymous sites to phylogenetic trees. This visualization makes it straightforward to compare rates of recent functional divergence within the context of rates of sequence divergence across the entire gene family. In this way, a researcher chould hone in on gene clades that are likely to exhibit high rates of functional redundancy or, conversely, gene clades where there is a greater likelihood that novel functions have emerged.

Previous research has found evidence of both broad genetic redundancy and specificity within closely related pairs or groups of IAA proteins [@overvoorde_functional_2005; @winkler_variation_2017]. For example, the *iaa8-1 iaa9-1* double mutant and the *iaa5-1 iaa6-1 iaa19-1* triple mutant have wild-type phenotypes [@overvoorde_functional_2005], yet the *IAA6/19* sister pair has significant differences in expression patterns, protein abundances and functions suggesting they have undergone functional specialization since their divergence [@winkler_variation_2017]. A closer examination of the *IAA19* and *IAA6* pair in *A. thaliana* and *A. lyrata* found that *IAA6* was diversifying far faster than [@winkler_variation_2017]. Consistent with these results, ViVa revealed higher conservation for *IAA19* (${\pi_N/\pi_S}$ = 0.55) compared to *IAA6* (${\pi_N/\pi_S}$ = 2.3) (Figure \@ref(fig:IAAtree)), and also detected high diversity within the same regions of *IAA6* as seen in [@winkler_variation_2017]. This pattern---one *Aux/IAA* sister showing high nonsynonymous diversity while the other sister was more conserved---was observed frequently across the family, suggesting this could be a recurring feature in *Aux/IAA* evolution supporting the large diversity in auxin functions.

@@

The notable exception to the patter of one conserved and one diersifying sister was observed within the Group B *IAA* genes (*IAA10*/*IAA11*, *IAA12*/*IAA13*, *IAA18*/*IAA26*, and *IAA20*/*IAA30*) [@remington_contrasting_2004]. Among these sister pairs only *IAA12* and *IAA13* have a noticeable difference in nonsynonymous diversity. In our analysis, *IAA10* and *IAA11* both showed functional conservation (${\pi_N/\pi_S}$ of 0.80 and 0.67 respectively). In support of *IAA10* and *IAA11* playing largely redundant roles, the *Arabidopsis thaliana* EFP browser indicates that *IAA10* and *IAA11* have almost identical expression patterns.

### 2018-11-21 manuscript/index.Rmd
Added the app figure to the results section. also added a sub-section "ViVa R Package"
updates from JLN 181118

abstract: Thousands of sequenced genomes are now publically available and represent a large swath of plant evolution; yet, much of this data remains inaccessible to researchers without significant bioinformatics experience. Here, we present a webtool called ViVa (Visualizing Variation) which aims to empower any researcher to take advantage of the amazing genetic resource collected in the Arabidopsis 1001 Genomes Project. ViVa facilates data mining on the gene, gene family or gene network level. To test the utility and accessibility of ViVa, we assembled a team with a range of expertise with biology and bioinformatics to analyze the natural variation in the well-studied nuclear auxin signaling pathway. Our analysis has provided further confirmation of existing knowledge and has also helped generate new hypotheses regarding this well studied pathway. These results highlight how natural variation can be used to generate and test hypotheses about less studied gene families and networks, especially when paired with biochemical and genetic characterization. ViVa is also readily extensible to newly emerging databases of genetic variation, such as the 3,000 Rice Genomes Project.

>>>>

abstract: Thousands of sequenced genomes are now publically available and represent a large swath of plant evolution; yet, much of this data remains inaccessible to researchers without significant bioinformatics experience. Here, we present a webtool called ViVa (Visualizing Variation) which aims to empower any researcher to take advantage of the amazing genetic resource collected in the Arabidopsis 1001 Genomes Project (http://1001genomes.org). ViVa facilates data mining on the gene, gene family or gene network level. To test the utility and accessibility of ViVa, we assembled a team with a range of expertise with biology and bioinformatics to analyze the natural variation in the well-studied nuclear auxin signaling pathway. Our analysis has provided further confirmation of existing knowledge and has also helped generate new hypotheses regarding this well studied pathway. These results highlight how natural variation could be used to generate and test hypotheses about less studied gene families and networks, especially when paired with biochemical and genetic characterization. ViVa is also readily extensible to newly emerging databases of genetic variation, such as the 3,000 Rice Genomes Project (http://snp-seek.irri.org/).
